### 2021-09-06 rad-icfp-2021/relwork.tex
starting fixing things after POPL reviews

\noindent \textbf{Array primitives}
We focused on giving reverse derivatives for a small set of array primitives. 
Other common primitives include filter, flatten, gather. 
These functions can easily be added to our Source language once we provide a reverse derivative for them.
This is reminiscent of hand-crafted derivatives present in large AD frameworks (usually hundreds), 
and these can already be added in added in our language. 
As shown in \ref{sub:lift_recipe}, one mostly only needs to make sure that the provided transpose Jacobian is correct.

>>>>

\noindent \textbf{Array primitives}
We focused on giving reverse derivatives for a small set of array primitives. 
Other common primitives include filter, flatten, gather. 
These functions can easily be added to our Source language once we provide a reverse derivative for them.
This is reminiscent of hand-crafted derivatives present in large AD frameworks (usually hundreds), 
and these can already be added in our language. 
As shown in \ref{sub:lift_recipe}, one mostly only needs to make sure that the provided transpose Jacobian is correct.

### 2021-09-06 rad-icfp-2021/unf.tex
saving more space
removed unfinished example
starting fixing things after POPL reviews

Following the intuition highlighted in Section~\ref{subsec:insights}, 
we present a new language which we call Unary Normal Form (UNF). 
There is a Source UNF for the Source language and a Target UNF for the Target language.
Just as Target is an extension of Source, Target UNF is an extension of Source UNF.
They serve as intermediate representation our the reverse-mode compilation pipeline (see Figure~\ref{fig:outline}).

@@

The Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar and type system of target UNF is given in Figure~\ref{fig:unf_target_grammar}.

@@

We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
This term in Source UNF carries the context of the original term in Source.
$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
Because of this, when we are trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
When we need to pass the result of type !A! of !e$_{1}$! through !e$_{2}$! which does not take it as an input, we can transform !e$_{2}$! to accept and pass it.
It's the equivalent in UNF of a weakening.
This weakening is required when we transform a non-unary operator such as $+$ or !pair!, as we transform the arguments in order and we need to keep all of their results.
We write this new term $(\widetilde{e_{2}})_{A}$ and it defined by induction on !e$_{2}$! as follows. 
We assume given a type !A! that needs to be passed, and drop the index !A!.

\begin{tabular}{r c l}
    $\widetilde{var_{T;i}}$  &=& $var_{T,A;i}$ \\
    $\widetilde{op_{T;n}}$  &=& $op_{T,A;n}$ \\§
    $\widetilde{pair_{T;B\times C}}$  &=& $pair_{T,A;B\times C}$ \\
    $\widetilde{proj_{T_1;T_2;T_3}}$  &=& $proj_{T_1;T_2;T_3,A}$ \\
    $\widetilde{e_1 \comp e_2}$  &=& $\widetilde{e_1}\comp\widetilde{e_2}$ \\
    $\widetilde{map2_{T;x,y.e}}$  &=& $map2_{T,A;x,y.e}$ \\
    $\widetilde{reduce_{T;x,y.e_1;e_2}}$ &=& $reduce_{T,A;x,y.e_1;e_2}$ \\
\end{tabular}

@@

\begin{example}
    Continuing again with our previous example !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$!, 
    we have, after removing a few unnecessary let-bindings:

    \begin{tabular}{l}
        !x$_1$,x$_2$,x$_3$:$\reals$,x$_4$:$\reals^3$->$\rho$ $\vdash$ $\UNFSymbol^{-1}$($\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$): $(\reals\times\reals) \times (\reals\times\reals\times(\reals\times\reals))$->$\rho$! \\
        = !TODO!
    \end{tabular}
\end{example}

>>>>

Following the intuition highlighted in Section~\ref{subsec:insights}, 
we present a new language which we call Unary Normal Form (UNF). 
There is a Source UNF for the Source language and a Target UNF for the Target language.
Just as Target is an extension of Source, Target UNF is an extension of Source UNF.
They serve as intermediate representations in the reverse-mode compilation pipeline (see Figure~\ref{fig:outline}).

@@

Target UNF is an extension of Source UNF. 
Its types are lists of types of Target. 
Higher-order types are needed for the continuation introduced by reverse-mode. 
In addition, it contains several new primitives which represent the transpose Jacobian of the primitives of Source UNF.
Given a type !T$\defeq$ [A$_1$,$\ldots$,A$_n$]!, we write \underline{T} for the type !A$_1$ $\times\ldots\times$ A$_n$! of Target. 
Target UNF has an internal composition $\icomp$. 
!e$_{1}$ $\icomp$ e$_{2}$! represents precomposition of a curried function !e$_{1}$! by a term !e$_{2}$!.
It is used by the reverse-mode transformation to pre-compose the continuation term by a transpose Jacobian.
The grammar and type system of target UNF is given in Figure~\ref{fig:unf_target_grammar}.

@@

We first give a translation from our Source to Source UNF in Figure~\ref{fig:source_to_unf}, which we call $\UNFSymbol$.
As discussed above, $\UNFSymbol$ sends a constant or an operator to a primitive in Source UNF. 
This term in Source UNF carries the context of the original term in Source.
$\UNFSymbol$ sequentializes a term, mimicking a left-to-right call-by-value evaluation.
Because of this, when we are trying to return a pair, this needs to be witnessed, and this is the role of !pair!.
When we need to pass the result of type !A! of !e$_{1}$! through !e$_{2}$! which does not take it as an input, we can transform !e$_{2}$! to accept and pass it.
It's the equivalent in UNF of a weakening.
This weakening is required when we transform a non-unary operator such as $+$ or !pair!, as we transform the arguments in order and we need to keep all of their results.
We write this new term $(\widetilde{e_{2}})_{A}$. It is defined by induction on !e$_{2}$! as follows. 
We assume given a type !A! that needs to be passed, and drop the index !A!.

\begin{tabular}{r c l}
    $\widetilde{var_{T;i}}$  &=& $var_{T,A;i}$ \\
    $\widetilde{op_{T;n}}$  &=& $op_{T,A;n}$ \\
    $\widetilde{pair_{T;B\times C}}$  &=& $pair_{T,A;B\times C}$ \\
    $\widetilde{proj_{T_1;T_2;T_3}}$  &=& $proj_{T_1;T_2;T_3,A}$ 
\end{tabular}
\begin{tabular}{r c l}
    $\widetilde{e_1 \comp e_2}$  &=& $\widetilde{e_1}\comp\widetilde{e_2}$ \\
    $\widetilde{map2_{T;x,y.e}}$  &=& $map2_{T,A;x,y.e}$ \\
    $\widetilde{reduce_{T;x,y.e_1;e_2}}$ &=& $reduce_{T,A;x,y.e_1;e_2}$
\end{tabular}

@@

% \begin{example}
%     Continuing again with our previous example !cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$!, 
%     we have, after removing a few unnecessary let-bindings:

%     \begin{tabular}{l}
%         !x$_1$,x$_2$,x$_3$:$\reals$,x$_4$:$\reals^3$->$\rho$ $\vdash$ $\UNFSymbol^{-1}$($\Dsynrevsymbol_{\rho}$(cos$_{\reals,\reals}$ $\comp$ pair$_{\reals,\reals; \reals\times\reals}$): $(\reals\times\reals) \times (\reals\times\reals\times(\reals\times\reals))$->$\rho$! \\
%         = !TODO!
%     \end{tabular}
% \end{example}
